---
title: "Demo for the R package paaPack with the NICU Studies"
author: |
  | Yan Li, Gen Li and Kun Chen
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: bookdown::html_document2
runtime: shiny
---

```{css, echo=FALSE}
p {
  font-size: 12pt;
}

.main-container {
  max-width: 1600px;
  margin-left: auto;
  margin-right: auto;
}
```


```{r setup, echo = FALSE, results = "hide", message = FALSE}
options(width = 100)
knitr::opts_chunk$set(fig.width = 16, fig.height = 9, dpi = 300,
                      out.width = "90%", fig.align = "center",
                      cache = TRUE)
ipak <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg))
    install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE, quietly = TRUE)
  invisible(NULL)
}
pkgs <- c("Rcpp", "ggplot2", "gridExtra", "bookdown", "pheatmap", "gplots", "ggforce", "shiny", "shinyWidgets")
ipak(pkgs)
```


# Overview

The **paaPack** is designed to perform hierarchical Principal Amalgamation Analysis (HPAA) with or without the 
guidance of taxonomic tree structure, and provide several useful graphical tools for visualizing the results of HPAA, 
including 1) hierarchical dendrograms to visualize the full path of amalgamations, 2) the scree plot showing the percentage change 
in the diversity loss along with the changes of number of compositions, and 3) the ordination plot showing the changes in the 
between-sample distance patterns before and after HPAA with any given number of principal compositions (PC). In this tutorial, we 
also provide a R shiny app to dynamically visualize the changes of ordination plots along with the path of HPAA, i.e., 
from the largest number to the smallest number of PCs.

## Set up

```{r install_package, echo = TRUE, eval= TRUE, results = "hide", message = FALSE}
install.packages("../Rpackage/paaPack_0.0-1.tar.gz", repos = NULL, type="source")
library(paaPack)
#### functions in paaPack
help(hPAA)  #### fit HPAA models
help(plotHPAA)  #### dendrogram showing the hierarchical amalgamation
help(plotLine)  #### scree plot showing the percentage
help(plotMDS)  #### ordination plot
```

```{r, echo=FALSE, results="hide", message = FALSE}
## load the package
library(paaPack)
```

The **paaPack** provides a main function `hPAA()` to perform the hierarchical Principal Amalgamation Analysis (HPAA). 
To use the function, the analyst should provide:

+ The compositional data with row representing compositions for each subject and 
  column recording the original components/taxa.
+ The taxonomic tree structure as a vector with dimension same as the number of taxa. The format of the taxonomic 
  vector is the same as output format of the commonly used bioinformatics data processing software **Mothur**. 
  That is, each element of the vector denotes the full taxonomic ranks from kingdom to genus, species level of the taxon, 
  with ranks separated by semicolon. For example a typical element of the taxonomic vector could be
  `k_Bacteria;p_Actinobacteria;c_Actinobacteria;o_Bifidobacteriales;f_Bifidobacteriaceae;g_Bifidobacterium;s_longum`.
  The taxonomic structure is optional. If taxonomy is not provided, the unconstrained HPAA without tree guidance is performed.
  
+ The diversity measures used in the HPAA analysis, and indicate whether strong or weak taxonomic hierarchy is applied in the analysis.

Then a set of plotting methods are provided taking the object of class "hPAA" from the `hPAA()` as input, including `plotHPAA()` 
for the dendrogram showing the full path of hierarchical amalgamation, `plotLine()` for the scree plot showing the percentage change in 
the diversity loss with the changing in the number of principal compositions, and `plotMDS()` for the ordination plot showing the changes 
in the between-sample distance patterns before and after HPAA. In each function, a group of graphical arguments for shaping the figures could be 
specified. For details, the analyst could refer to the documentations of the functions using `help()`.

# Illustration

In the following sections, we use the NICU data to illustrate the visualization tools provided by **paaPack**. The codes for constructing 
all the figures are documented in the corresponding code chunks of the source R markdown file. These tools can be extremely useful for 
visualizing and understanding compositional data, as well as helping to determine the desired number of principal compositions in practice.

```{r, echo=FALSE, results="hide", message = FALSE}
#### fit the models for different diversity measures
load("NICU.rda")
## reconstruct the taxonomic tree structure and add labels for the loweast taxonomic rank to the structure.
taxonomy[!grepl(pattern = "g__", taxonomy)] <-
  paste0(taxonomy[!grepl(pattern = "g__", taxonomy)],
         rownames(taxonomy)[!grepl(pattern = "g__", taxonomy)])
method.list <- c("Simpson", "Shannon", "Bray-Curtis")
for(method in method.list) {
  tmp <- new.env()
  tmp$datPlot.strong <- hPAA(compDat, method = method, taxonomy = taxonomy, strong = TRUE)  ## strong
  tmp$datPlot.weak <- hPAA(compDat, method = method, taxonomy = taxonomy, strong = FALSE)  ## weak
  tmp$datPlot.uc <- hPAA(compDat, method = method)  ## unconstraint
  tmp$datPlot.uc$taxonomy <- taxonomy
  eval(call("<-", as.name(method), tmp))
}
```


## Dendrogram

We construct a HPAA dendrogram to simultaneously visualize both the tree diagram of the successive amalgamations and the taxonomic 
structure of the taxon using the function `plotHPAA()`. To illustrate, Figure \@ref(fig:dendrogram) shows the HPAA dendrogram from 
performing HPAA with SDI loss and strong taxonomic hierarchy on the NICU data. The top part of figure shows the dendrogram of amalgamations, 
where the $y-$axis shows the percentage decrease in total diversity as measured by SDI (on the log-scale) along the successive amalgamations, 
from the bottom to the top. As such, any horizontal cut of the dendrogram at a desired level of diversity loss/preservation shows the 
corresponding amalgamated data. In particular, each red dashed horizontal line indicates the steps at which the original data are aggregated 
to a higher taxonomic rank. It shows that, for example, aggregating data to the order level (22 taxa or principal compositions left) through 
HPAA leads to 22.3\% loss in total SDI. At the bottom part, we use color bars to show taxonomic structure of the taxa, where in each horizontal 
bar taxa of the same color belong to the same category of that rank.

```{r dendrogram, echo=FALSE, message=FALSE, fig.width=16, fig.height=9, out.width = "70%", fig.cap="The NICU data: Dendrogram of HPAA with SDI and strong taxonomic hierarchy."}
par(mfrow = c(1, 1), mar = c(5.1, 4.1, 2.1, 0.5), cex.lab = 1.4)
plotHPAA(object = Simpson$datPlot.strong, taxonomy = TRUE, rank.line = TRUE,
         lab.cex = 1, 
         main = "HPAA dendrogram with strong taxonomic hierarchy",
         ylab = "Percentage change (%)",
         y_scale = 1.1, y_shift = -0.5, jitters = TRUE,
         cex.rowLabels = 1.3 ## rank labels
         )
```

Then we display the results under different ways of taxonomy guidance of each loss function in one combined figure for intuitive 
comparison. Figures \@ref(fig:dendro-simpson) and \@ref(fig:dendro-bray-curtis) show HPAA dendrograms with SDI loss and BC loss, 
respectively, under all three levels of taxonomy guidance. Not surprisingly, the patterns of amalgamations vary under different settings. 
Without taxonomic constrain, the change in diversity appears to be very smooth along the amalgamations, but the resulting principal 
compositions may not be easily interpretable, as indicated by the mixed color patterns in the color bars of the taxonomic rank. On the 
other hand, for the setting of strong taxonomic hierarchy, while the principal compositions are forced to closely follow the taxonomic 
structure, the percentage change in diversity tends to exhibit dramatic jumps, especially at the steps that the last remaining taxon at 
a lower taxonomic rank is forced to be aggregated to a higher rank. As a compromise, for the setting of weak taxonomic hierarchy, the 
resulting principal compositions remain interpretable, and the percentage change in diversity remains smooth and can be quite close to 
that of the unconstrained setting in the early stage of amalgamations.

```{r, echo = FALSE, message=FALSE}
datastr.list <- c("datPlot.uc", "datPlot.weak", "datPlot.strong")
plottitle.list <- 
  c("HPAA dendrogram (unconstrained)",
    "HPAA dendrogram with weak taxonomic hierarchy",
    "HPAA dendrogram with strong taxonomic hierarchy")
```

```{r dendro-simpson, echo=FALSE, message=FALSE, fig.width=23, fig.height=6, fig.cap="The NICU data: HPAA dendrograms with SDI and different constrains on taxonomic hierarchy."}
par(mfrow = c(1, 3))
for(i in 1:length(datastr.list)) {
  ## get the corresponding result of the selected taxonomy guidance
  datastr <- datastr.list[i]
  plottitle <- plottitle.list[i]
  plotData <- get(datastr, envir = Simpson)
  ## plot the dendrogram
  par(mar = c(5.1, 4.1, 2.1, 0.5), cex.lab = 1.4)
  plotHPAA(object = plotData, taxonomy = TRUE, rank.line = FALSE,
         main = plottitle,
         ylab = "Percentage change (%)",
         lab.cex = 1, y_scale = 1.1, y_shift = -0.45, jitters = TRUE,
         cex.rowLabels = 1.3 ## rank labels
         )
}
```

```{r dendro-bray-curtis, echo=FALSE, message=FALSE, fig.width=23, fig.height=6, fig.cap="The NICU data: HPAA dendrograms with Bray-Curtis and different constrains on taxonomic hierarchy."}
par(mfrow = c(1, 3))
for(i in 1:length(datastr.list)) {
  ## get the corresponding result of the selected taxonomy guidance
  datastr <- datastr.list[i]
  plottitle <- plottitle.list[i]
  plotData <- get(datastr, envir = `Bray-Curtis`)
  ## plot the dendrogram
  par(mar = c(5.1, 4.1, 2.1, 0.5), cex.lab = 1.4)
  plotHPAA(object = plotData, taxonomy = TRUE, rank.line = FALSE,
         main = plottitle,
         ylab = "Percentage change (%)",
         lab.cex = 1, y_scale = 1.1, y_shift = -0.6, jitters = TRUE,
         cex.rowLabels = 1.3 ## rank labels
         )
}
```


## Scree Plot

Next we use the function `plotLine()` to construct the scree plot for the results of HPAA under different types of taxonomy guidance. 
The scree plot shows the percentage change in the diversity loss as a function of the number of principal compositions. 
Figure \@ref(fig:percent-line) shows the scree plots from performing HPAA on the NICU data under different settings. The difference 
among the three levels of taxonomic guidance is very revealing, which confirms the previous observation from the dendrograms that 
the setting of weak taxonomic hierarchy reaches a good balance between preserving information and interpretability.

```{r percent-line, echo=FALSE, message=FALSE, fig.width=15, fig.height=5, fig.cap="The NICU data: Scree plots for HPAA (Percentage change in diversity vs. number of principal compositions)."}
layout(matrix(c(1, 2, 3, 4, 4, 4), ncol=3, byrow=TRUE), heights=c(4.5, 0.5))
par(mar=c(5.1,4.1,4.1,2.1),cex.main = 1.4, cex.lab = 1.4)
## the vector specifies the corresponding environment variable for the different methods
dat.list <- c("Simpson", "Shannon", "Bray-Curtis")
for(k in 1:3) {
  ## get the corresponding environment variable for the selected method
  dat <- get(dat.list[k])
  for(i in 1:length(datastr.list)) {
    ## get the results of different constraints, unconstrained, weak hierarchy, and strong hierarchy respectively.
    datastr <- datastr.list[i]
    plotData <- get(datastr, envir = dat)
    ## plot scree line.
    plotLine(plotData, col = i, xaxt = 'n', lty = i, pch = i)
    par(new= TRUE)
  }
  axis(1, at=1:length(plotData$height), labels=factor(length(plotData$height):1))
  par(new=FALSE)
}
par(mai=c(0.1,0,0,0), cex = 0.8)
plot.new()
legend(x="center", ncol=3,legend=c("Unconstrained","Weak hierarchy","Strong hierarchy"),
       lty = 1:3, col = 1:3, pch = 1:3)
```


## Ordination plot

Finally, we use `plotMDS()` to construct ordination plot to visualize the changes in the between-sample distance patterns before and 
after HPAA with any given number of principal compositions. Specifically, in the provided function we perform the non-metric multidimensional 
scaling (NMDS) analysis with Bray–Curtis dissimilarity on the combined original data and the principal compositions from HPAA, which 
produces a low-dimensional ordination plot of all samples before and after amalgamation. For each sample, it is represented by a pair 
of points from either the original data or the principal compositions; the smallest circle that covers the pair is drawn, whose radius 
then indicates the level of distortion due to HPAA data reduction. The ordination plots from performing HPAA on the NICU data with three different 
loss functions and weak taxonomic hierarchy are shown in Figure \@ref(fig:MDS), in which 20 principal compositions are kept (the number of PCs can be 
updated via the corresponding parameter of the plotMDS function). All three settings preserve the between-sample diversity reasonably well, 
as indicated by the fact that the circles generally have a small radius; as expected, HPAA with the BC loss performs the best as it directly targets on preserving between-sample diversity.

```{r MDS, echo=FALSE, message=FALSE, fig.width=14, fig.height=5.5, fig.cap="The NICU data: 2D NMDS ordination plots for comparing original and principal com- positions from HPAA with weak taxonomic hierarchy."}
dat.list <- c("Simpson", "Shannon", "Bray-Curtis")
for(k in 1:length(dat.list)) {
    ## get the corresponding environment variable for the selected method
  dat <- get(dat.list[k])
  ## select only the result of weak hierarchy according to the previous observations
  plotData <- get("datPlot.weak", envir = dat)
  ## set the cutoff and plot the MDS
  x <- paste("p", k, sep="")
  p <- plotMDS(plotData, cutoff = 20, size = 1.2)
  eval(call("<-", as.name(x), p))
}
get_legend<-function(myggplot){
  tmp <- ggplot_gtable(ggplot_build(myggplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)
}
## get the legend of the figure
legend <- get_legend(p1)
p1$theme$legend.position <- p2$theme$legend.position <- p3$theme$legend.position <- "none"
## combine all the figures together
grid.arrange(p1, p2, p3, legend, 
             heights=c(5, 1), 
             layout_matrix = rbind(c(1, 2, 3), cbind(4, 4, 4)))
```


## R Shiny App

In this section, we construct an interactive R Shiny app to combine the scree plots and the ordination plots described in
the previous section. The Shiny app allows users to view the comparison of amalgamated data and the orgininal data with different
number of principle components.

```{r, echo=FALSE, cache=FALSE}
## load the data for the Simpson's index under weak hierarchy
plotData <- get("datPlot.weak", envir = Simpson)
#### ui.R
ui <- fluidPage(
  titlePanel(paste("NMDS Plot: HPAA with ", plotData$diversity.method, " index")),  # Add a title panel
  sidebarLayout(
    sidebarPanel(
      sliderTextInput(inputId = "Cutoff", label = "Number of PCs",
                      choices = seq(from = ncol(plotData$compDat) - 1, to = 2),
                      selected = 40,  width = '100%'),
      width = 6
    ),
    mainPanel(
      column(6, align="center", plotOutput("myPercent", width = "80%", height = "600px")),
      column(6, align="center", plotOutput("myMDS", height = "600px")),
      width = 12
    )
  )
)

#### server.R
server <- function(input, output) {
  #### output: the MDS plot
  output$myMDS <- renderPlot({
    plotMDS(plotData, cutoff = input$Cutoff)
  })
  output$myPercent <- renderPlot({
    #### the percent change line plot
    plotLine(plotData, col = 1, lty = 1)
    pData <- paaPack:::pComp(plotData)
    #### number of PCs input$Cutoff
    #### mark the current step
    points(ncol(plotData$compDat) - input$Cutoff, pData$height[ncol(compDat) - input$Cutoff], col = "red", pch = 1)
    points(ncol(plotData$compDat) - input$Cutoff, pData$height[ncol(compDat) - input$Cutoff], col = "red", pch = "+")
    abline(v = ncol(plotData$compDat) - input$Cutoff, col = "red", lty = 2)
    abline(h = pData$height[ncol(plotData$compDat) - input$Cutoff], col = "red", lty = 2)
    text(-1, pData$height[ncol(plotData$compDat) - input$Cutoff] + 3 ,
         labels = sprintf("Percentage Change: %s%%",
                          round(pData$height[ncol(compDat) - input$Cutoff], 3)), pos = 4, col = "red")
  })
}

## Run the app
shinyApp(ui = ui, server = server,
  options = list(height = 1000))
```